---
title: "Gathering Carbon Pools Data"
output: html_document
---
```{r loading libraries}
library(neonUtilities)
library(tidyverse)
library(lubridate)
```
First you will find your data product’s unique ID.
Go to NEONscience.org and click the dropdown menu under “Data and Samples.” Click “Data Portal”, then “Explore Data Products.”

Use the search bar on the left hand side of the page to search for your data product
i.	  Trees: “Vegetation structure”
ii.	  Soil:  “Soil physical and chemical properties, distributed initial characterization”
iii.	Roots: “Root biomass and chemistry, periodic"

Find your data product and note the data product ID. It will begin with DP1.
Trees: _DP1.10098.001_
Soil:  _DP1.10047.001_
Roots: _DP1.10067.001_
```{r}
soils_data <- loadByProduct(dpID="DP1.10047.001", site= c("BART", "KONZ", "STEI", "WREF"), 
                            startdate= "2012-01", enddate= "2019-12", package= "basic")
                                 
roots_data <- loadByProduct(dpID="DP1.10067.001", site= c("BART", "KONZ", "STEI", "WREF"),
                            startdate="2012-01", enddate="2019-12", package= "basic")

trees_data <- loadByProduct(dpID="DP1.10098.001", site= c("BART", "KONZ", "STEI", "WREF"), 
                            startdate= "2012-01", enddate= "2019-12", package= "basic")

```

List to Global Environment:
```{r}
list2env(soils_data, .GlobalEnv)
list2env(roots_data, .GlobalEnv)
list2env(trees_data, .GlobalEnv)

```

Making csv for the dataframes:
```{r}
# Soil Data
write.csv(spc_biogeochem,  file= "soil_BioGeoChem.csv")
write.csv(spc_particlesize,file= "soil_ParticleSize.csv")
write.csv(spc_bulkdensity, file= "soil_BulkDensity.csv")
write.csv(variables_10047, file= "NEON Metadata Soil.csv")

# Root Data
write.csv(bbc_percore,  file= "bbc_percore_roots.csv")
write.csv(bbc_rootmass, file= "bbc_rootmass.csv")
write.csv(variables_10067, file= "NEON Metadata Roots.csv")

# Tree Data
read.csv(vst_apparentindividual, file= "vst_apparentindividual.csv")
read.csv(vst_perplotperyear, file= "vst_perplotperyear.csv")
read.csv(variables_10098, file= "NEON Metadata Trees.csv")
```
